Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-10 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
small formatting/quotes

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Funding Specialist Expertise Group Discussion}
\label{sec:appendix_funding_spec_expert}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed. And this brings up the flip side question, “Do scientists really want to be writing code?”

@@

An alternative approach has been the provision of programming expertise as
part of national compute services. The US XSEDE project's extended collaboration
service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
“permanent” staff.  When PIs request allocations on XSEDE resources, there is a 
finite pool of developer time that can be awarded, typically for one year only, 
and at partial effort, typically 20 percent or so. The finite time allowed provides
motivation for the scientist and their group to work closely the developer and to 
become educated in what the developer is doing, so they can sustain the effort, once 
the ECSS period is over. This funding mechanism can be highly efficient for scientific
problems, because the developer pool assembled by the research providers are, by definition, 
expert in the characteristics of their specific resource, and can very quickly assess the
scientists needs, and what it will take to implement software that meets the user’s needs. 
However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
program, it should not be relied upon as a long term solution to acquiring this type of capacity.

@@

The Canarie group (Canada) accepts proposals for providing services to broad communities, 
integrate people who are doing things that are complementary, the goal is to make the available 
stack more robust and richer for everyone. They offer short rounds of funding that can
have as a key metric creating some useful functionality that shows a diversity
of input and draws from across disciplines, then more funding could follow. This
could apply within or across institutions.

There can be problems communicating across cultural barriers, with domain scientists seeing developers 
as “other”. Collaboration, and tools to fund, encourage or motivate collaboration are
extremely important.

@@

The goal for Sloan is to provide success stories, to provide demonstrable value
in the presence of data scientists on university staff. These stories are the
basis for arguments to the host organization. This is an effort to create
awareness of the value of research scientist developers. Embedding with
scientists, and adding spare capacity is critical to making the innovation
possible. This model is essentially to argue for permanent budget lines to
support data scientists as part of university staff hires, just as with core
facilities. This could become a fee-for-service model requested by grant
funding, just as DNA sequencing is for core facilities, if it becomes apparent
that this gives competitive advantage to the University’s research effort.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Funding Research Programmer Expertise Group Discussion}
\label{sec:appendix_funding_spec_expert}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed. And this brings up the flip side question, ``Do scientists really want to be writing code?''

@@

An alternative approach has been the provision of programming expertise as
part of national compute services. The US XSEDE project's extended collaboration
service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
``permanent'' staff.  When PIs request allocations on XSEDE resources, there is a 
finite pool of developer time that can be awarded, typically for one year only, 
and at partial effort, typically 20 percent or so. The finite time allowed provides
motivation for the scientist and their group to work closely the developer and to 
become educated in what the developer is doing, so they can sustain the effort, once 
the ECSS period is over. This funding mechanism can be highly efficient for scientific
problems, because the developer pool assembled by the research providers are, by definition, 
expert in the characteristics of their specific resource, and can very quickly assess the
scientists needs, and what it will take to implement software that meets the user's needs. 
However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
program, it should not be relied upon as a long term solution to acquiring this type of capacity.

@@

The CANARIE group (Canada) accepts proposals for providing services to broad communities, 
integrate people who are doing things that are complementary, the goal is to make the available 
stack more robust and richer for everyone. They offer short rounds of funding that can
have as a key metric creating some useful functionality that shows a diversity
of input and draws from across disciplines, then more funding could follow. This
could apply within or across institutions.

There can be problems communicating across cultural barriers, with domain scientists seeing developers 
as ``other''. Collaboration, and tools to fund, encourage or motivate collaboration are
extremely important.

@@

The goal for Sloan is to provide success stories, to provide demonstrable value
in the presence of data scientists on university staff. These stories are the
basis for arguments to the host organization. This is an effort to create
awareness of the value of research scientist developers. Embedding with
scientists, and adding spare capacity is critical to making the innovation
possible. This model is essentially to argue for permanent budget lines to
support data scientists as part of university staff hires, just as with core
facilities. This could become a fee-for-service model requested by grant
funding, just as DNA sequencing is for core facilities, if it becomes apparent
that this gives competitive advantage to the University's research effort.

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex
small formatting/quotes

\katznote{I don't think this section title is exactly right - are there some other options?  Also, the title of Appendix~\ref{sec:appendix_funding_spec_expert} should match, whatever is chosen.  Maybe `Funding Research Programmer Expertise'?}

@@

One solution is to fund dedicated software engineering roles for major research software projects at national laboratories
or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
However, this strategy has limited application, as much of the body of software is created and maintained in research
universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
traditionally supported university cyberinfrastructure, library organizations, or research offices. Alternatively, these
groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
will vary from institution to institution, but the mandate and ways of working should be similar.

@@

The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSE's who do not identify themselves as such at this time. We will collect information as to the organizational models under which these groups function, and how they are funded. For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not. Are these developers paid by the University or through a program supported by research grants/individual PIs? How did they bootstrap the developer track to get this started? How successful is the university in getting investigators to pay for fractional RSEs?  We will author a report describing our findings, should funding be available to conduct the investigation.

>>>>

%\katznote{I don't think this section title is exactly right - are there some other options?  Also, the title of Appendix~\ref{sec:appendix_funding_spec_expert} should match, whatever is chosen.  Maybe `Funding Research Programmer Expertise'?}

@@

One solution is to fund dedicated software engineering roles for major research software projects at national laboratories
or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
However, this strategy has limited application, as much of the body of software is created and maintained in research
universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group.
This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
traditionally supported university cyberinfrastructure, library organizations, or research offices. Alternatively, these
groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
will vary from institution to institution, but the mandate and ways of working should be similar.

@@

The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSEs who do not identify themselves as such at this time. We will collect information as to the organizational models under which these groups function, and how they are funded. For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not. Are these developers paid by the University or through a program supported by research grants/individual PIs? How did they bootstrap the developer track to get this started? How successful is the university in getting investigators to pay for fractional RSEs?  We will author a report describing our findings, should funding be available to conduct the investigation.
